```{r}

library(tidyverse)
library(dobson)
library(dplyr)
library(GGally)

```

```{r}

tuesdata <- tidytuesdayR::tt_load('2022-04-12')

fuel_gdp <- tuesdata$fuel_gdp #fuel_gdp
death_source <- tuesdata$death_source #death_source

```

```{r}

pollution_deaths <- merge(fuel_gdp, death_source, by=c('Entity', 'Year', 'Code'))
pollution_deaths <- pollution_deaths %>% rename('Clean_Fuel_Access_Percent' = 'Access to clean fuels and technologies for cooking (% of population)',  
                                                'GDP_Per_Capita' = 'GDP per capita, PPP (constant 2017 international $)' ,
                                                'Population'= 'Population (historical estimates)', 
                                                'Deaths_Per_100000' = 'Deaths - Cause: All causes - Risk: Household air pollution from solid fuels - Sex: Both - Age: Age-standardized (Rate)', 
                                                'Country' = 'Entity')
pollution_deaths <- subset(pollution_deaths, select = -c(Continent))

#deaths is per 100000
pollution_deaths <- pollution_deaths[!is.na(pollution_deaths$Code),]
pollution_deaths <- na.omit(pollution_deaths)

```

```{r}

#scale GDP by thousands
#scale population by millions
#truncate decimal portion of deaths per 100000
pollution_deaths <- pollution_deaths %>% mutate(Population = Population/1000000, GDP_Per_Capita = GDP_Per_Capita/1000, Deaths_Per_100000 = as.integer(Deaths_Per_100000))
pollution_deaths <- subset(pollution_deaths, select = -Code)

# View(pollution_deaths)

```

```{r}

library(MASS)

null = glm(Deaths_Per_100000 ~ 1, family="poisson", data=pollution_deaths)
null

full = glm(Deaths_Per_100000 ~ Year * Clean_Fuel_Access_Percent * 
             GDP_Per_Capita * Population, data=pollution_deaths, 
           family="poisson")
full

partial = glm(Deaths_Per_100000 ~ Year + Clean_Fuel_Access_Percent + 
                GDP_Per_Capita + Population + Year:Clean_Fuel_Access_Percent + 
                Year:GDP_Per_Capita + Year:Population + Clean_Fuel_Access_Percent:GDP_Per_Capita +
                Clean_Fuel_Access_Percent:Population + GDP_Per_Capita:Population, data=pollution_deaths, 
              family="poisson")
partial

```

```{r}

pollution_deaths <- pollution_deaths %>% 
  mutate(Deaths = as.integer(Deaths_Per_100000>0))
View(pollution_deaths)

#fit bernoulli model 
#null
bern_null = glm(Deaths ~ 1, family="binomial", data=pollution_deaths)
#full
bern_full = glm(Deaths ~ Year * Clean_Fuel_Access_Percent * 
                  GDP_Per_Capita * Population, data=pollution_deaths, 
                family="binomial")


summary(bern_null)
summary(bern_full)

```

```{r}

n = length(pollution_deaths$Deaths)
stepAIC(bern_null, scope=list(upper=bern_full), direction="both", k=log(n))

#Best model so far - stepwise BIC on a Bernoulli model of deaths
model_final = glm(formula = Deaths ~ Clean_Fuel_Access_Percent + GDP_Per_Capita + 
                    Year + Clean_Fuel_Access_Percent:Year, family = "binomial", 
                  data = pollution_deaths)
summary(model_final)

```

```{r}

#fitting model from stepAIC
bern_model = glm(formula = Deaths ~ Clean_Fuel_Access_Percent + GDP_Per_Capita + 
                   Year + Clean_Fuel_Access_Percent:Year + Clean_Fuel_Access_Percent:GDP_Per_Capita, 
                 family = "binomial", data = pollution_deaths)
#dispersion check - looks fine
bern_quasi = glm(formula = Deaths ~ Clean_Fuel_Access_Percent + GDP_Per_Capita + 
                   Year + Clean_Fuel_Access_Percent:Year + Clean_Fuel_Access_Percent:GDP_Per_Capita, 
                 family = "quasibinomial", data = pollution_deaths)
summary(bern_model)
summary(bern_quasi)

```


Start of EDA Process
----------------------

From stepAIC, the model below is what was found to be the best fit model for
our data.

```{r}

model_final = glm(formula = Deaths ~ Clean_Fuel_Access_Percent + GDP_Per_Capita + 
                    Year + Clean_Fuel_Access_Percent:Year, family = "binomial", 
                  data = pollution_deaths)
model_final

```

First, we can ask how do these variables play on each other. Is there any
correlation amongst these variables and how linear are they to one another.

```{r}

# Pairs plot to show correlation between the 5 variables
ggpairs(model_final, columns = c(1,2,3,4), upper = list(continuous = wrap("cor", size = 5))) +
  theme(strip.text.x = element_text(size = 8),
        strip.text.y = element_text(size = 5.5))
ggsave("plots/eda-1.pdf")

```

Looking at each pair of variables, almost every pair has no correlation
whatsoever. Three pairs do stick out though that we can look into further.

$\bullet$ Clean_Fuel_Percent and Deaths_Per_100k (-0.553)
$\bullet$ GDP_Per_Capita and Deaths_Per_100k (-0.739)
$\bullet$ GDP_Per_Capita and Clean_Fuel_Percent (0.657)

Only 3 of the variables have some sort of correlation between them while the
other two (Year and Population) have almost zero correlation to any other
variable, which could be expected.

```{r}

# Each of the variables plotted against each other to see closer how they relate
# Correlation value is in the parenthesis
pdf("plots/eda-2.pdf")
par(mfrow = c(1, 3))
plot(pollution_deaths$Clean_Fuel_Access_Percent, pollution_deaths$Deaths, xlab = "Clean Fuel Access %", ylab = "Deaths", main = "Clean Fuel vs. Deaths ~ (-0.553)")
plot(pollution_deaths$GDP_Per_Capita, pollution_deaths$Deaths, xlab = "GDP per Capita", ylab = "Deaths", main = "GDP vs. Deaths ~ (-0.739)")
plot(pollution_deaths$Clean_Fuel_Access_Percent, pollution_deaths$GDP_Per_Capita, xlab = " Clean Fuel Access %", ylab = "GDP per Capita", main = "Clean Fuel vs. GDP ~ (0.657)")
dev.off()

```

From the pairs of variables with correlation, two have negative and one has positive.
With deaths being a binomial variable, the plots 1 and 2 are a little hard to read.

From plot 1,
looking away from Deaths = 1 and paying more attention to Deaths = 0, we see
that they become visible at 60%+ and become more abundant as clean fuel access
increases. Deaths will happen regardless and with the bold line at Deaths = 1 not
much can be taken from it.

From plot 2,.
where we relate GDP and Deaths, the correlation factor is a little better. It
being -0.739, it means that as deaths increase, GDP in said country will decrease,
which is logical. Unlike plot 1, we can look at Deaths = 1 to analyze instead of 0.
Deaths = 1 seem to vanish as GDP per Capita hits around 50, helping our logic
make sense.

For plot 3,
both variables are normal and therefore we have a normal plot. With this one being
the only positive correlation out of the three, it should make sense that
as clean fuel becomes more available, the GDP of the country should increase.
As for  Clean Fuel >= 90, GDP of countries skyrocket, maybe proving all countries
should strive towards those numbers.

Moving away from the pairs plots and more at the residuals of the data. Is there
anything here that raise any questions?

```{r}

#get residuals
model_resid <- resid(model_final, type="deviance")

#residuals vs. fitted looks fine
pdf("plots/eda-3.pdf")
plot(model_final$fitted.values, model_resid, xlab = "Fitted Values", ylab = "Residuals")
dev.off()

#issues - more high fuel access percent in data, fanning in residuals there
pdf("plots/eda-4.pdf")
hist(pollution_deaths$Clean_Fuel_Access_Percent, xlab = "Clean Fuel %", main="Histogram of Clean Fuel Access %")
dev.off()
pdf("plots/eda-5.pdf")
plot(pollution_deaths$Clean_Fuel_Access_Percent, model_resid, xlab = "Clean Fuel %",
     ylab = "Residuals")
dev.off()

#issues - more high gdp per capita in data, fanning in residuals there
pdf("plots/eda-6.pdf")
hist(pollution_deaths$GDP_Per_Capita, xlab= "GDP per Capita", main="Histogram of GDP Per Capita")
dev.off()
pdf("plots/eda-7.pdf")
plot(pollution_deaths$GDP_Per_Capita, model_resid, xlab= "GDP per Capita", ylab = "Residuals")
dev.off()

```

From simply looking at the residual models against Clean Fuel and GDP, the
relationship is good when Clean Fuel is closer and 0 and GDP closer to 100+.

There is a little flutter at the end of Clean Fuel and the data trends up. What
could be the reason for that? From the histogram, most countries reside in that
area and they all have the same trend.

The same can be said about the GDP plots where most countries reside at GDP <50.

With the focus being on Clean Fuel and GDP, we can pull plot 3 from the pairs plots
to try and help us.

```{r}
pdf("plots/eda-8.pdf")
plot(pollution_deaths$Clean_Fuel_Access_Percent, pollution_deaths$GDP_Per_Capita, xlab = " Clean Fuel Access %", ylab = "GDP per Capita", main = "Clean Fuel vs. GDP ~ (0.657)")
dev.off()

```

What is interesting is the reverse is happening in the residual plots vs. Clean Fuel
and GDP. But with the plot above, there is a positive correlation. 

The outliers seen in each of these plots are certain countries that have their
data skewed in some way, which is why there are trends within the major trend.
The hump in the plot above relates to one country, Equatorial Guinea. So, in the
clusters outside of the main picture are not multiple countries but just one. Guinea
will be looked into later.

```{r}

final_fit <- model_final$fitted.values
model_resid <- resid(model_final, type="deviance")
final_df <- data.frame(final_fit, model_resid)

ggplot(data = final_df, aes(final_fit, model_resid)) +
  geom_jitter(height = 1/10, width = 0.01, size = 3, alpha = 1/4) +
  geom_smooth() +
  geom_hline(yintercept = 0, col = 2) +
  labs(title = "Deviance residuals")
ggsave("plots/eda-9.pdf")

```

With our new plot, we see the relationship between the two lines. What is good
is the abundance of points are towards the zero line. But, both taper off when at
the extremes.

```{r}

count <- 1:3044
df_resid <- data.frame(count, model_resid)

resid_pos <- df_resid %>%
  filter(model_resid > 0)
nrow(resid_pos)

resid_neg <- df_resid %>%
  filter(model_resid < 0)
nrow(resid_neg)

poll_pos <- pollution_deaths %>%
  filter(Deaths > 0)
nrow(poll_pos)

```

Deaths = 1 is the top line while Deaths = 0 is the bottom.

```{r}
pdf("plots/eda-10.pdf")
binned_fitted <- cut(model_final$fitted.values, breaks = 20)
plot(binned_fitted)
dev.off()
table(binned_fitted)

```

From the abundance each line sees on the "head", Deaths = 1 sees 1710 occurrences
while Deaths = 0 head is only 171, a tenth of the size. An helpful indicator of this
is in the bias of the blue line being dragged once it gets all the way to the right.

Since we know that the residuals are bunched up in terms of Deaths = 1 and Deaths = 0,
we can use that to look back on the Clean Fuel vs Residual and GDP vs Residual
to see that when Death = 0, the residuals for Clean Fuel are far from 0 and not good,
while trending to 0 as Clean Fuel also increases. For GDP, a little more detail
work is needed before making an assumption.

```{r}

gdp_pos_75 <- df_resid %>%
  filter(model_resid > 0) %>%
  filter(model_resid <= 0.75)

gdp_pos_50 <- df_resid %>%
  filter(model_resid > 0) %>%
  filter(model_resid <= 0.5)

num_pos_75 <- nrow(gdp_pos_75)
num_pos_50 <- nrow(gdp_pos_50)
den_pos_75 <- nrow(poll_pos)
den_pos_50 <- nrow(poll_pos)
num_pos_75/den_pos_75*100
num_pos_50/den_pos_50*100

```

88.52% of the data for Deaths = 1 appear when residuals are 0 < residual < 0.75 and
81.02$ show when residuals are 0 < residual < 0.50. Meaning, the vast majority are
towards the residual line while from afar is might seem like the data is more skewed
than it is.

With Deaths = 1, is might make sense to see that when GDP is closer to 0, more
deaths are present, showing a better residual and closer to 0 as well. As GDP
increases, that fluttering of the data appears and that residual value rises.

With Deaths = 0, the same assumption can be made. Residuals are bad when Deaths = 0
when GDP is low but as GDP increases, the residuals come very close to 0 because
more money relates to a better clean fuel access (as seen in the pos. corr plot),
resulting in lower death numbers.

